---
title: "Reproducible Research: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
---
## Data Processing
We begin by loading libraries and setting a few global parameters:
```{r}
  library(knitr)
  #opts_chunk$set(echo=TRUE)       ## set global parameter for echo
  setwd("~/datasciencecoursera/RepResProj2/")
```

We first download and unzip the data (if necessary):
```{r}
  #Download file if it does not exist

  if (!file.exists("repdata-data-StormData.csv.bz2")) {
      message("Downloading data...")
      fileURL <- "http://bit.ly/1uNSAQY"
      zipfile = "repdata-data-StormData.csv.bz2"
      download.file(fileURL, destfile=zipfile, method="curl")
  }
```

We then read the data into R
```{r cache=TRUE}
  # Load the csv file and assign it to a variable
  file    = "repdata-data-StormData.csv.bz2"
  raw    =  read.csv(file,stringsAsFactors = FALSE)
```

## Synopsis
The purpose of the analysis is to determine which types of events are most harmful with respect to population health in the United States.

## Results

To begin the analysis, we explore the overall layout of the data:
```{r}
dim(raw)
names(raw)
```

End line. + 2
